---
title: "Part 4: Factors"
subtitle: Introduction to R
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
urlcolor: blue
---

```{r include=FALSE}
knitr::opts_chunk$set(eval=FALSE, include=TRUE)
```

This assignment consists of six parts:

- Intro to Basics
- Vectors
- Matrices
- Factors (this document)
- Data frames
- Lists

Create a script called `hw02-4.R` and save it in your `hw02` folder.

After you complete each exercise, commit and push your R script to your remote repo. See [Part 0](./00_intro.html) for instructions. Do *not* push this document. 

***

## 4.1 What is a factor?

In R, a **factor** is vector data used for categorical variables, also called discrete data. Discrete data can belong only to one of a limited number of categories, such as plant, animal, fungus, or bacteria. The Linnean classification hierarchy used by biologists is another example of categorical data.  A species can belong to only one genus, one family, one order, and so on. Factors are different from continuous data that can have an unlimited number of possible values. R handles factors and continuous variables differently as you will learn during this course.

To create factors in R, you first create a vector with the names of your discrete categories. You then pass that vector to the `factor()` function. Study this example.

```{r eval=TRUE}
# Genotype can be homozygous or heterozyous. This vector contains
# the genotype of six individuals.
genotype_vector <- c("Heterozygote", "Heterozygote", "Homozygote", "Heterozygote", "Homozygote", "Heterozygote")

genotype_factor <- factor(genotype_vector)

genotype_factor
```

The output lists the elements of the vector and also **Levels.** Levels are the unique categories that `factor()` found in the vector. 


#### Instructions

For this exercise, you will assume that you counted the number and species of bumble bees (genus *Bombus*) that visitied a single flowering plant in one hour.  Your notes show the order of the visitors:

- kirbiellus
- kirbiellus 
- flavifrons 
- kirbiellus 
- bifarius
- flavifrons 
- kirbiellus

Add code to your script to

- Make a character vector called `bee_visitors_vector` with these seven visitors. Your vector must reflect the visiting order listed above.

- Use `factor()` to make a factor from `bee_visitors_vector`. Assign the new factor to `bee_visitors_factor` 

- Print out `bee_visitors_factor` to be sure you have the order and three levels. A incorrect number of levels suggests you might have spelled a name incorrectly.

```{r}
# Make a character vector called `bee_visitors_vector`


# Make a `bee_visitors_factor` factor from `bee_visitors_vector`. 


# Print out bee_visitors_factor


#
```

Your results should look like this. Notice that the levels are arranged alphabetically. R does this automatically, which may not be what you want. You'll learn later how to arrange factors in the order you want.

```{r answer4.1, eval=TRUE, include=TRUE, echo=FALSE}
# Make a character vector called `bee_visitors_vector`
bee_visitors_vector <- c("kirbiellus", "kirbiellus", "flavifrons", "kirbiellus", "bifarius", "flavifrons", "kirbiellus")

# Make a `bee_visitors_factor` factor from `bee_visitors_vector`. 
bee_visitors_factor <- factor(bee_visitors_vector)

# Print out bee_visitors_factor
bee_visitors_factor
```



## 4.2 Factor types

Categorical variables can be **ordinal** or **nominal.** Ordinal categories are ordered. For example, the bee species listed above have proboscis lengths that can be categorized as short, intermediate, or long An order is implied because short < intermediate < long lengths. You can choose to order the categories from short to long or long to short. What's important is that an order is implied. 

Nominal categories (nomen = name) do not have an implied order. The three species of bees are nominal categories. There is no implied order to the names of the three species. *Bombus kirbiellus* is not any more or less a bumble bee than *B. flavifrons* or *B. bifarius.* They're just different.

Study the example below. The *Bacillus* factor is nominal because no order is implied for the bacteria. `factor()` defaults to alphabetical order for the levels. The [allozyme](https://en.wikipedia.org/wiki/Alloenzyme) factor is ordinal because allozymes move through an electrophoretic gel at different speeds (here, slow, medium, fast). Notice that the order of the levels is shown as `Slow < Medium < Fast`.

```{r eval=TRUE}
# Bacillus bacteria
bacillus_vector <- c("subtilus", "cereus", "pasteurii", "sphaericus")
bacillus_factor <- factor(bacillus_vector)
bacillus_factor

# Allozymes are slow, medium, or fast
allozyme_vector <- c("Medium", "Slow", "Fast", "Slow", "Medium")
allozyme_factor <- factor(allozyme_vector, order = TRUE, levels = c("Slow", "Medium", "Fast"))
allozyme_factor

```

Pay attention to the arguments for `factor()` in the allozyme example. In addition to the vector, the `order = TRUE` argument tells the function that order matters. The `levels = c("Slow", "Medium", "Fast")` argument tells the function the specific order of the levels. 

```{r}
allozyme_vector <- c("Medium", "Slow", "Fast", "Slow", "Medium")
allozyme_factor <- factor(allozyme_vector, order = TRUE, levels = c("Slow", "Medium", "Fast"))
```

How would you rewrite the call to `factor()` to arrange the allozyme levels from fast to slow?


#### Instructions

Here are the proboscis lengths for the three species of *Bombus* you surveyed.

- kirbiellus: Long
- flavifrons: Intermediate
- bifarius: Short

Add code to your script to

- Create a vector of seven lengths that matches the order of the bees in your nominal vector above (the order you observed them). Name that vector `proboscis_vector`.

- Create an ordinal factor called `proboscis_factor`, with the lengths ordered from long to short.

- Display `proboscis factor` to be sure you have it correct.

**Hint:** Remember that case matters. 


```{r}
# Create `proboscis_vector` with seven elements


# Create an ordered `proboscis_factor` ordered from long to short.


# Display the contents of `proboscis_factor


#
```

```{r eval=TRUE}
# Create `proboscis_vector` with seven elements
proboscis_vector <- c("Long", "Long", "Intermediate", "Long", "Short", "Intermediate", "Long")

# Create an ordered `proboscis_factor` ordered from long to short.
proboscis_factor <- factor(proboscis_vector, ordered = TRUE, levels = c("Long", "Intermediate", "Short"))

# Display the contents of `proboscis_factor
proboscis_factor

#
```

## 4.3 Adding or changing factor levels

What if you have nominal data that is unordered but you decide an order would be appropriate for your analysis? What if you want to change the order of ordinal data? The `levels()` function allows you to change the order of a factor without redefining the factor.  

For example, say that you were studying photoperiod (light availability) on flowering time in [*Asclepias tuberosa.*](https://www.fs.fed.us/wildflowers/plant-of-the-week/asclepias_tuberosa.shtml) You sampled plants from Louisiana, Arkansas, Missouri, Iowa, and Wisconsin.  At first, you think a state is a state so you categorize the data as nominal; order doesn't matter.  Then, you realize that the more northerly a state, the shorter the photoperiod. Study this code.

```{r eval=TRUE}
## Original nominal factor
state_vector <- c("Louisiana", "Arkansas", "Missouri", "Iowa", "Wisconsin")
state_factor <- factor(state_vector)

# Levels defaults to alphabetical order
state_factor


# Use levels to reorder the states from south to north.
levels(state_factor) <- c("Louisiana", "Arkansas", "Missouri", "Iowa", "Wisconsin")

# The state now appear in latitudinal order from south to north
state_factor
```

#### Instructions

Add code (and comments, right?) to your script to 

- Use `levels()` to change proboscis length order from short to long.

*Be careful:* If you do not specify the new order as a vector, then `levels()` changes the factor to unordered nominal. This is useful if you need to change an ordinal factor to nominal but if you are changing order, then be sure you provide the new order as a vector.


## 4.4 Summarizing a factor

A useful function that you will use often is `summary()` which provides a summary of the information contained in a variable. For example, `summary(allozyme_factor)` summarizes how many observations of each level are stored in the variable.

```{r eval=TRUE, echo=FALSE}
summary(allozyme_factor)
```

#### Instructions

Add code to your script to 

- Use the `summary()` function on the `bee_visitors_factor` and the `proboscis_factor`. You do not have to save the results to a variable.



## 4.6 Battle of the sexes

You might wonder what happens when you try to compare elements of a factor. In `factor_survey_vector` you have a factor with two levels: `"Male"` and `"Female"`. But how does R value these relative to each other?

#### Instructions

- Read the code in the editor and then run it to test if `male` is greater than (`>`) `female`.

```{r}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")

# Male
male <- factor_survey_vector[1]

# Female
female <- factor_survey_vector[2]

# Battle of the sexes: Male 'larger' than female?
male > female
```

## 4.7 Ordered factors

Since `"Male"` and `"Female"` are unordered (or nominal) factor levels, R returns a warning message, telling you that the greater than operator is not meaningful. As seen before, R attaches an equal value to the levels for such factors.

But this is not always the case! Sometimes you will also deal with factors that do have a natural ordering between its categories. If this is the case, we have to make sure that we pass this information to R.

Let us say that you are leading a research team of five data analysts and that you want to evaluate their performance. To do this, you track their speed, evaluate each analyst as `"slow"`, `"medium"` or `"fast"`, and save the results in `speed_vector`.

#### Instructions

As a first step, assign `speed_vector` a vector with 5 entries, one for each analyst. Each entry should be either `"slow"`, `"medium"`, or `"fast"`. Use the list below:

- Analyst 1 is medium,
- Analyst 2 is slow,
- Analyst 3 is slow,
- Analyst 4 is medium and
- Analyst 5 is fast.

No need to specify these are factors yet.

```{r}
# Create speed_vector


#
```

```{r answer4.7, include=FALSE}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
```

## 4.8 Ordered factors (2)

`speed_vector` should be converted to an ordinal factor since its categories have a natural ordering. By default, the function `factor()` transforms `speed_vector` into an unordered factor. To create an ordered factor, you have to add two additional arguments: `ordered` and `levels`.

```{r}
factor(some_vector,
       ordered = TRUE,
       levels = c("lev1", "lev2" ...))
```

By setting the argument `ordered` to `TRUE` in the function `factor()`, you indicate that the factor is ordered. With the argument `levels` you give the values of the factor in the correct order.

#### Instructions

- From `speed_vector`, create an ordered factor vector: `factor_speed_vector`. Set `ordered` to `TRUE`, and set `levels` to `c("slow", "medium", "fast")`.
- Print out `factor_speed_vector`
- Use `summary()` to summarize `factor_speed_vector`

```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector


# Print and summarize factor_speed_vector


#
```

```{r answer4.8, include=FALSE}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Print and summarize factor_speed_vector
factor_speed_vector
summary(factor_speed_vector)
```

## 4.9 Comparing ordered factors

Having a bad day at work, 'data analyst number two' enters your office and starts complaining that 'data analyst number five' is slowing down the entire project. Since you know that 'data analyst number two' has the reputation of being a smarty-pants, you first decide to check if his statement is true.

The fact that `factor_speed_vector` is now ordered enables us to compare different elements (the data analysts in this case). You can simply do this by using the well-known operators.

#### Instructions

- Use `[2]` to select from `factor_speed_vector` the factor value for the second data analyst. Store it as `da2`.

- Use `[5]` to select the `factor_speed_vector` factor value for the fifth data analyst. Store it as `da5`.

- Check if `da2` is greater than `da5`; simply print out the result. Remember that you can use the `>` operator to check whether one element is larger than the other.

```{r}
# Create factor_speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Factor value for second data analyst


# Factor value for fifth data analyst


# Is data analyst 2 faster than data analyst 5?


#
```

```{r answer4.9, include=FALSE}
# Create factor_speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Factor value for second data analyst
da2 <- factor_speed_vector[2]

# Factor value for fifth data analyst
da5 <- factor_speed_vector[5]

# Is data analyst 2 faster than data analyst 5?
2 > 5
```


