---
title: "4: Factors"
subtitle: Introduction to R
output: html_notebook
editor_options:
  chunk_output_type: inline
---

## 4.1 What's a factor and why would you use it?

#### 4.1 Instructions

- Assign to variable `theory` the value "factors for categorical variables".

```{r}
# Assign to the variable theory what this chapter is about!


#
```

## 4.2 What's a factor and why would you use it? (2)

#### Instructions

- Convert the character vector `sex_vector` to a factor with `factor()` and assign the result to `factor_sex_vector`

- Print out `factor_sex_vector` and assert that R prints out the factor levels below the actual values.

```{r}
# Sex vector
sex_vector <- c("Male", "Female", "Female", "Male", "Male")

# Convert sex_vector to a factor


# Print out factor_sex_vector


#
```


## 4.3 What's a factor and why would you use it? (3)

#### Instructions

Run the script below to check how R constructs and prints nominal and ordinal variables. Do not worry if you do not understand all the code just yet, we will get to that.


```{r}
# Animals
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector

# Temperature
temperature_vector <- c("High", "Low", "High","Low", "Medium")
factor_temperature_vector <- factor(temperature_vector, order = TRUE, levels = c("Low", "Medium", "High"))
factor_temperature_vector
```

## 4.4 Factor levels

#### Instructions

- Check out the code that builds a factor vector from `survey_vector`. You should use `factor_survey_vector` in the next instruction.

- Change the factor levels of `factor_survey_vector` to `c("Female", "Male")`. Mind the order of the vector elements here.

```{r}
# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <- c("F", "M")

factor_survey_vector
```


## 4.5 Summarizing a factor

#### Instructions

- Ask for a `summary()` of the `survey_vector` and `factor_survey_vector`. Interpret the results of both vectors. Are they both equally useful in this case?

```{r}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")

# Generate summary for survey_vector


# Generate summary for factor_survey_vector


#
```

## 4.6 Battle of the sexes

#### Instructions

- Read the code in the editor and then run it to test if `male` is greater than (`>`) `female`.

```{r}
# Build factor_survey_vector with clean levels
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")

# Male
male <- factor_survey_vector[1]

# Female
female <- factor_survey_vector[2]

# Battle of the sexes: Male 'larger' than female?
male > female
```

## 4.7 Ordered factors

#### Instructions

As a first step, assign `speed_vector` a vector with 5 entries, one for each analyst. Each entry should be either `"slow"`, `"medium"`, or `"fast"`. Use the list below:

- Analyst 1 is medium,
- Analyst 2 is slow,
- Analyst 3 is slow,
- Analyst 4 is medium and
- Analyst 5 is fast.

No need to specify these are factors yet.

```{r}
# Create speed_vector


#
```

## 4.8 Ordered factors (2)

#### Instructions

- From `speed_vector`, create an ordered factor vector: `factor_speed_vector`. Set `ordered` to `TRUE`, and set `levels` to `c("slow", "medium", "fast")`.
- Print out `factor_speed_vector`
- Use `summary()` to summarize `factor_speed_vector`

```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector


# Print and summarize factor_speed_vector


#
```

## 4.9 Comparing ordered factors

#### Instructions

- Use `[2]` to select from `factor_speed_vector` the factor value for the second data analyst. Store it as `da2`.

- Use `[5]` to select the `factor_speed_vector` factor value for the fifth data analyst. Store it as `da5`.

- Check if `da2` is greater than `da5`; simply print out the result. Remember that you can use the `>` operator to check whether one element is larger than the other.

```{r}
# Create factor_speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")
factor_speed_vector <- factor(speed_vector, ordered = TRUE, levels = c("slow", "medium", "fast"))

# Factor value for second data analyst


# Factor value for fifth data analyst


# Is data analyst 2 faster than data analyst 5?


#
```

