---
title: "HW 11: Data Visualization I (con't)"
subtitle: "Graphical Analysis of Biological Data"
output:
  pdf_document: default
  html_document:
    df_print: paged
    toc: yes
#    latex_engine: lualatex
urlcolor: blue
---

```{r include=FALSE}
knitr::opts_chunk$set(include = TRUE, eval = TRUE, message=FALSE)
```


By the end of this assignment, you should demonstrate an ability to visually analyze data for 

- outliers,
- normality, 
- relationships among variables, and
- interactions.

Click on any blue text to visit the external website.

**Note:** If you contact me for help or (better yet) open an issue in the [public discussion forum,](https://github.com/SEMO-GABD/public_discussion) please include the code that is not working and also tell me what you have tried.

## Preparation

- Open your `.Rproj` project file in RStudio.

- Create an `hw11` folder inside the same folder as your project file.

- Create a notebook file with the usual `<lastname>_hw11.Rmd` name and usual YAML header, and save it in your `hw11` folder. 

- Right-click and save [aegla_crabs.csv](https://raw.githubusercontent.com/SEMO-GABD/semo-gabd.github.io/master/assignments/hw11/aegla_crabs.csv) to your data folder.

- Add a code chunk to load the `tidyverse`, `here`, `ggally`, and `patchwork` libraries.

- Commit early. Commit often.  Push regularly but at least push your completed assignment.

- Download and open [A protocol for data exploration to avoid common statistical problems](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x) (free access) by Zuur et al. 2010. 


```{r}
library(tidyverse)
library(here)
library(GGally)
library(patchwork)
```

***

## Habitat parameters for *Aegla* crabs

### Data exploration: crabs

*Aegla* is a genus of South American freshwater crabs. Satterlee et al., as part of his research, collected water quality parameters. Those data are included in [aegla_crabs.csv](https://raw.githubusercontent.com/SEMO-GABD/semo-gabd.github.io/master/assignments/hw11/aegla_crabs.csv). 

The variables are

- Site (sampling locations)

- Width, Depth (stream width and depth)

- Flow (rate of water movement)

- AT, WT (air and water temperature)

- pH (pH; What did you expect me to say, log molar hydrogen ion concentration?)

- TDS (total dissolved solids)

- Cond (conductivity)

- N, Ni, Nt (Nitrogen, Nitrate, Nitrite concentrations)

- Phos, Mg, Ca (phosphate, magnesium, calcium concentrations)

- Crab, Shrimp, Aeglam, AeglaFe, AeglaFo (number of shrimp and crabs sampled)

Your goal is simple but it will take time.

- Import the raw data from `aegla_crabs.csv`.

- Use `select()` to remove the Site, AT, WT, Crab, Shrimp, Aeglam, AeglaFe, AeglaFo columns.

- Use `drop_na()` to remove the one row with missing water quality data.

- Create a `row_order` dummy variable like we did for the sparrows. 

- You must use the pipe (`%>%`) to write efficient code.

- **Explore the data.** There are three apparent outliers among the variables but only one that I think is an actual outlier. Find them and justify your choices.

- **Describe the results** that you see in your figures. Use the notebook to describe your interpretations of the data. Your descriptions must be more than single sentences although they do not have to be lengthy paragraphs. Two or three sentences is probably sufficient for most cases. I am looking for evidence that you studied and thought about the data.  *I want you to think like the scientist you are becoming.*

- Choose four variables to make Cleveland plots, and make a 2x2 grid with the four plots, using `patchwork`.

- Choose three different variables to make histograms. Play with the `bins` or `binwidths` argument that you think reveals the data well. *Hint:* the default `bins = 30` that `geom_histogram` is usually not a suitable choice. Save each histogram to a unique variable.

- Use the same three variables to produce density plots. Save each plot to a unique variable.

- Use `patchwork to produce a 2 column matrix with the histograms in the left column and the corresponding density plots in the right column.

- Use `ggpairs` to explore the relationships among all the variables. 

The *Aegla* crabs data is not suitable for testing for interactions. I'll see if I can find an opportunity to do that in a future assignment. I know you are thrilled.

***

*et Vóila*
